Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-06-23 manuscript.tex
Changes from first read-through (need to expand conclusions)

\begin{abstract}
  A depletion solver has been implemented in OpenMC and is described herein. The
  depletion solver is implemented in Python and interfaces with OpenMC's
  transport solver through a C++ API, which enables an in-memory
  depletion--transport coupling. Multiple integration methods for advancing in
  time have been implemented and exhibit tradeoffs in cost, accuracy, and memory
  use. For all time integration methods, evaluation of the matrix exponential is
  performed using the incomplete partial fraction form of the Chebyshev rational
  approximation method.

@@

With continued advances in computing, Monte Carlo (MC) particle transport
simulations are now widely used in reactor design and analysis in a variety of
roles. The earliest efforts to couple MC particle transport with depletion date
back to the 1990s~\citep{moore1995inel,trellue1998lanl}. These early efforts
involved taking existing MC and depletion codes, such as
MCNP~\citep{goorley2012nt} and ORIGEN~\citep{croff1983nt}, and putting a
``wrapper'' around them that orchestrated the transfer of solutions from one
code to the other. Thus, the actual transport/depletion codes themselves
required no source code changes. In the last decade, many MC code developers
have chosen to directly include a depletion solver as a first-rate feature,
including MCNPX~\citep{waters2007aip} (which has since been merged into
MCNP6~\citep{goorley2012nt}), Serpent~\citep{leppanen2015ane}, and
MC21~\citep{griesheimer2015ane}.

In the present work, we describe a new depletion solver in
OpenMC~\citep{romano2015ane1}, an open source, community-developed MC particle
transport code. To date, OpenMC has not had a built-in depletion solver. This
has led to numerous efforts in the community toward developing coupled
transport-depletion calculations with OpenMC~\citep{gul2017ane,
lanversin2017icone, lanversin2019phd, liu2019nst, zhuang2020pne, zhao2020ned,
zhao2020cpc, zhang2020ane}. While these efforts have proven the viability of
performing such coupled calculations with OpenMC, none of them has resulted in a
validated, openly-available, and generalized capability that OpenMC users can
take advantage of. The capability that is described herein is part of the OpenMC
project and can be found in version 0.11~\citep{openmc-0110} of the code or
later.

@@

Although the neutron transport equation is time-dependent, the timescale over
which material compositions change is sufficiently long such that the transport
equation can be solved as a time-independent equation. Without loss of
generality, let us write the burnup matrix solely as a function of $\vect{n}$
so that \cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\vect{n}}{dt} = \vect{A}(\vect{n})\vect{n}, \quad \vect{n}(0) =
  \vect{n}_0.
\end{equation}
This better reflects the fact that the time, $t$, is generally not an input to
the transport equation, and hence the burnup matrix is uniquely determined by
the value of $\vect{n}$.

@@

A series of \emph{predictor--corrector} methods that use multiple stages offer
improved accuracy over the predictor method. The simplest of these methods is
the CE/CM algorithm, which is the default algorithm used in
MCNP6~\citep{fensin2006tans}. In this method, $\vect{A}$ is evaluated at the
beginning of the step and used to deplete the composition over half the step.
Then, $\vect{A}$ is reevaluated using the predicted middle-of-step composition
and used to deplete over the full step. The algorithm can be expressed as
\begin{equation}
  \begin{split}
    \vect{n}_{i+1/2} &= \exp \left (\frac{h}{2}\vect{A}(\vect{n}_i) \right) \vect{n}_i \\
    \vect{n}_{i+1} &= \exp \left(h \vect{A}(\vect{n}_{i+1/2}) \right) \vect{n}_i.
  \end{split}
\end{equation}
This method requires two transport solutions and two matrix exponential solves
per timestep (twice as expensive as the predictor method) and is second-order
accurate.

@@

The Magnus expansion can be used as the basis for developing numerical
integration algorithms and has been widely studied. The typical approach is to
truncate the $\vect{\Omega}$ series at an appropriate order and apply an
approximation to the multivariate integrals in \cref{eq:magnus-terms}. In
particular, OpenMC uses a fourth-order commutator-free\footnote{The integrator
is called commutator-free because it does not require computing any matrix
commutators like those that appear in \cref{eq:magnus-terms}. Other methods
based on the Magnus expansion do require computing matrix commutators, some of
which were explored by \citet{josey2017phd}.} exponential
integrator~\citep{blanes2006anm} that is derived from the Magnus expansion in
order to solve \cref{eq:linear-ode}. In this approximation, the Magnus expansion
is truncated at two terms and then expressed as the composition of two
exponentials. For a system of ODEs in the form of \cref{eq:linear-ode}, the
resulting method is
\begin{equation}
  \label{eq:compose4}
  \vect{n}_{i+1} = \exp (a_2 \vect{A}_1 + a_1 \vect{A}_2)
  \exp (a_1 \vect{A}_1 + a_2 \vect{A}_2) \vect{n}_i
\end{equation}
where $\vect{A}_1$ and $\vect{A}_2$ correspond $\hat{\vect{A}}$ evaluated at the
nodes of a two-point Gauss-Legendre quadrature over the interval $[0,h]$:
\begin{equation}
  \begin{split}
    a_1 &= \frac{1}{4} + \frac{\sqrt{3}}{6}, \; a_2 = \frac{1}{4} - \frac{\sqrt{3}}{6} \\
    c_1 &= \frac{1}{2} - \frac{\sqrt{3}}{6}, \; c_2 = \frac{1}{2} + \frac{\sqrt{3}}{6} \\
    \vect{A}_1 &= h\hat{\vect{A}}\left ( c_1 h\right ), \; \vect{A}_2 = h\hat{\vect{A}}\left ( c_2 h\right ).
  \end{split}
\end{equation}
Evaluating $\vect{A}_1$ and $\vect{A}_2$ using \cref{eq:linear-interp} and
substituting in \cref{eq:compose4}, we obtain a CE/LI method based on the
fourth-order commutator-free Magnus integrator:
\begin{equation}
  \label{eq:celi-cfq4}
  \begin{split}
  \vect{n}_{i+1}^p &= \exp \left ( h \vect{A}(\vect{n}_i ) \right ) \\
  \vect{n}_{i+1} &= \exp \left( \frac{h}{12} \vect{A}(\vect{n}_i) +
    \frac{5h}{12} \vect{A}(\vect{n}_{i+1}^p) \right)
  \exp \left( \frac{5h}{12} \vect{A}(\vect{n}_i) +
  \frac{h}{12} \vect{A}(\vect{n}_{i+1}^p) \right) \vect{n}_i.
  \end{split}
\end{equation}
\Cref{eq:celi-cfq4} requires two transport solutions and two matrix
exponentials, whereas the CE/LI method with substeps, \cref{eq:celi-substeps},
requires two transport solutions and $m$ matrix exponentials. A comparison of
these two approaches has shown that \cref{eq:celi-cfq4} achieves better accuracy
than the use of five substeps~\citep{josey2017phd}. OpenMC also includes an
integrator based on the LE/QI (linear extrapolation, quadratic interpolation)
method~\citep{isotalo2011ane2} where the integration of \cref{eq:linear-ode} (in
which case $\hat{\vect{A}}(t)$ is represented by a quadratic polynomial in $t$)
is performed using the fourth-order commutator-free Magnus integrator based on
\cref{eq:compose4}.

@@

All of the integration methods discussed thus far are explicit methods that are
only conditionally stable. There are many well-documented cases where
oscillatory behavior and diverging solutions have been observed. To address
this, \citet{dufek2013ane} proposed an implicit-Euler-like method that is
unconditionally stable. The main idea is to replace $\vect{A}(\vect{n}_i)$ with
$\vect{A}(\vect{n}_{i+1})$ in \cref{eq:predictor}:
\begin{equation}
  \label{eq:implicit}
  \vect{n}_{i=1} = \exp \left( h\vect{A}(\vect{n}_{i+1}) \right) \vect{n}_i.
\end{equation}
This results in a non-linear equation for $\vect{n}_{i+1}$ that can't be solved
directly. However, it is possible to solve for $\vect{n}_{i+1}$ using a
fixed-point iteration. \citet{dufek2013ane} utilized a stochastic approximation
based on the Robbins-Monro algorithm~\citep{robbins1951ams} in order to
guarantee convergence of the fixed-point iteration while accounting for
stochasticity in the MC transport solver.

@@

As we saw in \cref{sec:time_integration}, numerically integrating
\cref{eq:depletion-matrix} requires evaluating one or more matrix exponentials.
OpenMC uses the Chebyshev rational approximation method (CRAM), which was
introduced in a series of papers by Pusa~\citep{pusa2010nse,pusa2011nse}, to
evaluate matrix exponentials. In particular, OpenMC utilizes an incomplete
partial fraction (IPF) form~\citep{pusa2016nse} of CRAM that provides a good
balance of numerical stability and efficiency. In this representation the matrix
exponential is approximated as
\begin{equation}
    \exp(\vect{A}t) \approx \alpha_0 \prod\limits_{\ell=1}^{k/2} \left (
    \vect{I} + 2 \text{Re} \left ( \widetilde{\alpha}_\ell \left (\vect{A}t
    - \theta_\ell \vect{I} \right )^{-1} \right ) \right )
\end{equation}
where $k$ is the order of the approximation and $\alpha_0$,
$\widetilde{\alpha}_\ell$, and $\theta_\ell$ are coefficients that have been
tabulated for orders up to $k=48$. Rather than computing the full approximation
and then multiplying it by a vector, \cref{alg:cram}\footnote{The original
description of the algorithm presented by \citet{pusa2016nse} contains a typo.}
is used to incrementally apply the terms within the product. The $k$th order
approximation for CRAM requires solving $k/2$ sparse linear systems. OpenMC
relies on functionality from scipy.sparse.linalg for solving the linear system.
\begin{algorithm}[H]
  \caption{Incomplete partial fraction form of CRAM.}
  \label{alg:cram}
  \begin{algorithmic}[1]
    \State $\vect{n} \gets \vect{n_0}$
    \For{$\ell \gets 1$ \textbf{to} $k/2$}
      \State $\vect{n} \gets \vect{n} + 2\text{Re}(\widetilde{\alpha}_\ell
        (\vect{A}t - \theta_\ell)^{-1})\vect{n}$
    \EndFor
    \State $\vect{n} \gets \alpha_0 \vect{n}$
  \end{algorithmic}
\end{algorithm}

@@

In OpenMC, the burnup matrix is constructed based on data inside of a
\emph{depletion chain} file, which includes fundamental data gathered from ENDF
incident neutron, decay, and fission product yield (FPY) sublibraries. For each
nuclide, this file includes:
\begin{itemize}
  \item What transmutation reactions are possible, their Q values, and their products;
  \item If a nuclide is not stable, what decay modes are possible, their
  branching ratios, and their products; and
  \item If a nuclide is fissionable, the fission products yields at any number
  of incident neutron energies.
\end{itemize}

Depletion chains can be automatically created in the Python API in the following
manner. First, the user supplies a list of ENDF incident neutron, decay, and FPY
sublibrary files. The incident neutron files are used to determine what nuclear
reactions result in transmutation and what the corresponding Q values of the
reactions are. The decay files are used to determine the half-life and possible
decay modes of nuclides that are unstable. The FPY sublibraries are used to
determine what fission products are produced when a fission reaction occurs.
This information is then aggregated and written to an XML file that serves as
the depletion chain file that is used at runtime to construct a burnup matrix.

@@

Some nuclides that have fission reactions in ENDF/B-VII.1, such as $^{243}$Pu,
do not have corresponding FPYs. For these nuclides, a decision has to be made on
what set of FPY should be used when fission occurs. This issue was briefly
studied by \citet{fensin2015pne}, who found that the proportion of fission
reactions that occur in these nuclides is usually small enough that the choice
of FPYs is not of great importance. For the depletion chains used in OpenMC, we
have adopted the method used in Serpent, which is as follows:
\begin{enumerate}
  \item If the nuclide missing FPYs is a metastable state, use FPYs from the
  ground state;
  \item Use FPYs from an isotone if available;
  \item Use FPYs from $^{235}$U.
\end{enumerate}

@@

Like capture branching ratios, fission product yields (FPYs) also depend on the
incident neutron energy. ENDF fission product yield sublibraries typically
include yields tabulated at two or three energies (thermal, fast, and fusion
energies). Complicating matters is the fact that while these data are specified
for a single energy, they actually represent the observed FPYs for a spectrum of
energies (characterized by a single point). For example, FPYs are often given at
\SI{0.0253}{\electronvolt}, but this actually represents the observed FPY under
a thermal neutron spectrum that contains a continuum of energies. The ENDF-6
formats manual~\citep{trkov2018bnl} recommends using linear-linear interpolation
between energy points. However, at the end of a transport solve, we typically
have a one-group or multigroup fission reaction rate---it's not clear how one
should use this information to calculate an effective FPY.

@@

OpenMC's core particle transport solver is written in C++ and is built as a
single executable. Most of the user-facing code, however, is written in Python.
Generally, to run OpenMC, a user writes a Python script that utilizes classes
and functions defined as part of an application programming interface (API) to
generate XML input files that are then read by the executable. On top of this,
OpenMC is also built as a shared library with a C++ API that can be called by
external programs. OpenMC's Python API includes bindings to the C++ API; this
enables a single run to be broken up into individual components. For example, a
function call from the Python bindings to the C++ API can initialize the solver
(reading inputs, loading cross sections, etc.) without actually performing a
simulation. A separate function call can run the transport simulation. Further
function calls allow a power user to obtain information about tally results
directly in memory from the C++ API.

@@

The abstract \texttt{Integrator} class represents a time integration method.
This class effectively orchestrates the coupled transport-depletion simulation.
Subclasses of \texttt{Integrator} are required to implement a method that takes
an initial vector of nuclide densities, reaction rates returned from an
operator, a time interval, and a power, and returns the resulting nuclide
densities at the end of the step. Thus, \texttt{Integrator} utilizes
\texttt{TransportOperator} in order to obtain reaction rates given $\vect{n}$,
and once the burnup matrix is constructed, it utilizes \texttt{DepSystemSolver}
to deplete with a constant $\vect{A}$ as many times as is necessary. One example
of a subclass of \texttt{Integrator} is \texttt{CECMIntegrator}, which
implements the CE/CM method.

@@

Six coupled transport-depletion simulations were carried out with OpenMC to test
a variety of combinations of depletion chains and methods for handling FPY
energy dependence:
\begin{enumerate}
  \item Full depletion chain, thermal FPYs
  \item Full depletion chain, interpolated FPYs
  \item Reduced depletion chain, thermal FPYs
  \item CASL depletion chain, thermal FPYs
  %\item CASL depletion chain, FPYs based on cutoff method
  \item CASL depletion chain, interpolated FPYs
\end{enumerate}
Because Serpent doesn't have an option for changing the depletion chain or the
method for handling FPY energy dependence, a single coupled transport-depletion
simulation with Serpent was carried out. For all simulations, 20 inactive
batches, 200 active batches, and \num{4.0e6} particles per batch were used,
which resulted in an uncertainty on the effective multiplication factor of about
3--4 pcm.

@@

Four coupled transport-depletion simulations were carried out with OpenMC for
the SFR assembly model:
\begin{enumerate}
  \item Full depletion chain, interpolated FPYs, probability tables on
  \item CASL depletion chain, interpolated FPYs, probability tables on
  \item Full depletion chain, interpolated FPYs, probability tables off
  \item CASL depletion chain, interpolated FPYs, probability tables off
\end{enumerate}
The results for the PWR pincell model indicated that the reduced depletion chain
results in identical behavior to the full depletion chain, so we have not
included a case for the SFR assembly using the reduced depletion chain. Two
coupled transport-depletion simulation with Serpent was carried out, one with
probabilty tables turned on and one with them turned off. For all simulations,
20 inactive batches, 200 active batches, and \num{4.0e6} particles per batch
were used, which resulted in an uncertainty on the effective multiplication
factor of about 3--4 pcm.

@@

A new depletion solver has been implemented in the OpenMC Monte Carlo particle
transport code. The depletion solver is implemented primarily in Python 3 and
interfaces with OpenMC's transport solver through Python bindings to an
underlying C++ API, which enables data transfers to happen in memory rather than
through filesystem I/O. The design of the depletion solver relies on a set of
abstract classes that provide modularity and separation between difference
components, including the time integration methods, the matrix exponential
solver, and the transport operator. In principle, this modularity would allow
the depletion system to be used with another transport code. Rather than relying
on a single time integration method, OpenMC implements several time integration
schemes including the predictor method, standard predictor-corrector methods,
stochastic implicit Euler integrators, and several novel
methods~\citep{josey2017phd, josey2017jcp}. To evaluate the matrix exponential,
OpenMC relies on the incomplete partial fraction form of CRAM by default.

@@

>>>>

\begin{abstract}
  A depletion solver has been implemented in OpenMC and is described herein. The
  depletion solver is implemented in Python and interfaces with OpenMC's
  transport solver through a C++ application programming interface, which
  enables an in-memory transport--depletion coupling. Multiple integration
  methods for advancing in time have been implemented and exhibit tradeoffs in
  cost, accuracy, and memory use. For all time integration methods, evaluation
  of the matrix exponential is performed using the incomplete partial fraction
  form of the Chebyshev rational approximation method.

@@

Monte Carlo (MC) particle transport simulations are now widely used in reactor
design and analysis in a variety of roles thanks to continued advances in
computing. The earliest efforts to couple MC particle transport with depletion
date back to the 1990s~\citep{moore1995inel,trellue1998lanl}. These early
efforts involved taking existing MC and depletion codes, such as
MCNP~\citep{goorley2012nt} and ORIGEN~\citep{croff1983nt}, and putting a
``wrapper'' around them that orchestrated the transfer of solutions from one
code to the other. Thus, the actual transport/depletion codes themselves
required no source code changes. In the last decade, many MC code developers
have chosen to directly include a depletion solver as a first-rate feature,
including MCNPX~\citep{waters2007aip} (which has since been merged into
MCNP6~\citep{goorley2012nt}), Serpent~\citep{leppanen2015ane}, and
MC21~\citep{griesheimer2015ane}.

In the present work, we describe a new depletion solver in OpenMC, an open
source, community-developed MC particle transport code~\citep{romano2015ane1}.
To date, OpenMC has not had a built-in depletion solver. This has led to
numerous efforts in the community toward developing coupled transport--depletion
calculations with OpenMC~\citep{gul2017ane, lanversin2017icone,
lanversin2019phd, liu2019nst, zhuang2020pne, zhao2020ned, zhao2020cpc,
zhang2020ane}. While these efforts have proven the viability of performing such
coupled calculations with OpenMC, none of them has resulted in a validated,
openly-available, and generalized capability that OpenMC users can take
advantage of. The capability that is described herein is part of the OpenMC
project and can be found in version 0.11~\citep{openmc-0110} of the code or
later.

@@

Although the neutron transport equation is time-dependent, the timescale over
which material compositions change is sufficiently long such that the transport
equation can be solved as a steady state equation. Without loss of generality,
let us write the burnup matrix solely as a function of $\vect{n}$ so that
\cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\vect{n}}{dt} = \vect{A}(\vect{n})\vect{n}, \quad \vect{n}(0) =
  \vect{n}_0.
\end{equation}
This better reflects the fact that the time, $t$, is generally not an input to
the transport equation, and hence the burnup matrix is uniquely determined by
the value of $\vect{n}$.

@@

A series of \emph{predictor--corrector} methods that use multiple stages offer
improved accuracy over the predictor method. The simplest of these methods is
the CE/CM (constant extrapolation, constant midpoint) algorithm, which is the
default algorithm used in MCNP6~\citep{fensin2006tans}. In this method,
$\vect{A}$ is evaluated at the beginning of the step and used to deplete the
composition over half the step. Then, $\vect{A}$ is reevaluated using the
predicted middle-of-step composition and used to deplete over the full step. The
algorithm can be expressed as
\begin{equation}
  \begin{split}
    \vect{n}_{i+1/2} &= \exp \left (\frac{h}{2}\vect{A}(\vect{n}_i) \right) \vect{n}_i \\
    \vect{n}_{i+1} &= \exp \left(h \vect{A}(\vect{n}_{i+1/2}) \right) \vect{n}_i.
  \end{split}
\end{equation}
This method requires two transport solutions and two matrix exponential solves
per timestep (twice as expensive as the predictor method) and is second-order
accurate.

@@

The Magnus expansion can be used as the basis for developing numerical
integration algorithms and has been widely studied. The typical approach is to
truncate the $\vect{\Omega}$ series at an appropriate order and apply an
approximation to the multivariate integrals in \cref{eq:magnus-terms}. In
particular, OpenMC uses a fourth-order commutator-free\footnote{The integrator
is called commutator-free because it does not require computing any matrix
commutators like those that appear in \cref{eq:magnus-terms}. Other methods
based on the Magnus expansion do require computing matrix commutators, some of
which were explored by \citet{josey2017phd}.} exponential
integrator~\citep{blanes2006anm} that is derived from the Magnus expansion in
order to solve \cref{eq:linear-ode}. In this approximation, the Magnus expansion
is truncated at two terms and then expressed as the composition of two
exponentials. For a system of ODEs in the form of \cref{eq:linear-ode}, the
resulting method is
\begin{equation}
  \label{eq:compose4}
  \vect{n}_{i+1} = \exp (a_2 \vect{A}_1 + a_1 \vect{A}_2)
  \exp (a_1 \vect{A}_1 + a_2 \vect{A}_2) \vect{n}_i
\end{equation}
where $\vect{A}_1$ and $\vect{A}_2$ correspond to $\hat{\vect{A}}$ evaluated at
the nodes of a two-point Gauss-Legendre quadrature over the interval $[0,h]$:
\begin{equation}
  \begin{split}
    a_1 &= \frac{1}{4} + \frac{\sqrt{3}}{6}, \; a_2 = \frac{1}{4} - \frac{\sqrt{3}}{6} \\
    c_1 &= \frac{1}{2} - \frac{\sqrt{3}}{6}, \; c_2 = \frac{1}{2} + \frac{\sqrt{3}}{6} \\
    \vect{A}_1 &= h\hat{\vect{A}}\left ( c_1 h\right ), \; \vect{A}_2 = h\hat{\vect{A}}\left ( c_2 h\right ).
  \end{split}
\end{equation}
Evaluating $\vect{A}_1$ and $\vect{A}_2$ using \cref{eq:linear-interp} and
substituting in \cref{eq:compose4}, we obtain a CE/LI method based on the
fourth-order commutator-free Magnus integrator:
\begin{equation}
  \label{eq:celi-cfq4}
  \begin{split}
  \vect{n}_{i+1}^p &= \exp \left ( h \vect{A}(\vect{n}_i ) \right ) \\
  \vect{n}_{i+1} &= \exp \left( \frac{h}{12} \vect{A}(\vect{n}_i) +
    \frac{5h}{12} \vect{A}(\vect{n}_{i+1}^p) \right)
  \exp \left( \frac{5h}{12} \vect{A}(\vect{n}_i) +
  \frac{h}{12} \vect{A}(\vect{n}_{i+1}^p) \right) \vect{n}_i.
  \end{split}
\end{equation}
\Cref{eq:celi-cfq4} requires two transport solutions and two matrix
exponentials, whereas the CE/LI method with substeps, \cref{eq:celi-substeps},
requires two transport solutions and $m$ matrix exponentials. A comparison of
these approaches has shown that \cref{eq:celi-cfq4} achieves better accuracy
than the use of five substeps~\citep{josey2017phd}. OpenMC also includes an
integrator based on the LE/QI (linear extrapolation, quadratic interpolation)
method~\citep{isotalo2011ane2} where the integration of \cref{eq:linear-ode} (in
which case $\hat{\vect{A}}(t)$ is represented by a quadratic polynomial in $t$)
is performed using the fourth-order commutator-free Magnus integrator based on
\cref{eq:compose4}.

@@

All of the integration methods discussed thus far are explicit methods that are
only conditionally stable. There are many well-documented cases where
oscillatory behavior and diverging solutions have been observed. To address
this, \citet{dufek2013ane} proposed an implicit-Euler-like method that is
unconditionally stable. The main idea is to replace $\vect{A}(\vect{n}_i)$ with
$\vect{A}(\vect{n}_{i+1})$ in \cref{eq:predictor}:
\begin{equation}
  \label{eq:implicit}
  \vect{n}_{i=1} = \exp \left( h\vect{A}(\vect{n}_{i+1}) \right) \vect{n}_i.
\end{equation}
This results in a non-linear equation for $\vect{n}_{i+1}$ that cannot be solved
directly. However, it is possible to solve for $\vect{n}_{i+1}$ using a
fixed-point iteration. \citet{dufek2013ane} utilized a stochastic approximation
based on the Robbins--Monro algorithm~\citep{robbins1951ams} in order to
guarantee convergence of the fixed-point iteration while accounting for
stochasticity in the MC transport solver.

@@

As we saw in \cref{sec:time_integration}, numerically integrating
\cref{eq:depletion-matrix} requires evaluating one or more matrix exponentials.
OpenMC uses the Chebyshev rational approximation method (CRAM), which was
introduced in a series of papers by Pusa~\citep{pusa2010nse,pusa2011nse}, to
evaluate matrix exponentials. In particular, OpenMC utilizes an incomplete
partial fraction (IPF) form~\citep{pusa2016nse} of CRAM that provides a good
balance of numerical stability and efficiency. In this representation the matrix
exponential is approximated as
\begin{equation}
    \exp(\vect{A}t) \approx \alpha_0 \prod\limits_{\ell=1}^{k/2} \left (
    \vect{I} + 2 \text{Re} \left ( \widetilde{\alpha}_\ell \left (\vect{A}t
    - \theta_\ell \vect{I} \right )^{-1} \right ) \right )
\end{equation}
where $k$ is the order of the approximation and $\alpha_0$,
$\widetilde{\alpha}_\ell$, and $\theta_\ell$ are coefficients that have been
tabulated for orders up to $k=48$. Rather than computing the full approximation
and then multiplying it by a vector, \cref{alg:cram}\footnote{The original
description of the algorithm presented by \citet{pusa2016nse} contains a typo.}
is used to incrementally apply the terms within the product. The $k$th order
approximation for CRAM requires solving $k/2$ sparse linear systems. OpenMC
relies on functionality from scipy.sparse.linalg~\citep{virtanen2020nm} for
solving the linear system.
\begin{algorithm}[H]
  \caption{Incomplete partial fraction form of CRAM.}
  \label{alg:cram}
  \begin{algorithmic}[1]
    \State $\vect{n} \gets \vect{n_0}$
    \For{$\ell \gets 1$ \textbf{to} $k/2$}
      \State $\vect{n} \gets \vect{n} + 2\text{Re}(\widetilde{\alpha}_\ell
        (\vect{A}t - \theta_\ell)^{-1})\vect{n}$
    \EndFor
    \State $\vect{n} \gets \alpha_0 \vect{n}$
  \end{algorithmic}
\end{algorithm}

@@

In OpenMC, the burnup matrix is constructed based on data inside of a
\emph{depletion chain} file that includes fundamental data gathered from ENDF
incident neutron, decay, and fission product yield (FPY) sublibraries. For each
nuclide, this file includes:
\begin{itemize}
  \item What transmutation reactions are possible, their Q values, and their products;
  \item If a nuclide is not stable, what decay modes are possible, their
  branching ratios, and their products; and
  \item If a nuclide is fissionable, the fission products yields at any number
  of incident neutron energies.
\end{itemize}

Depletion chains can be automatically created in the Python application
programming interface (API) in the following manner. First, the user supplies a
list of ENDF incident neutron, decay, and FPY sublibrary files. The incident
neutron files are used to determine what nuclear reactions result in
transmutation and what the corresponding Q values of the reactions are. The
decay files are used to determine the half-life and possible decay modes of
nuclides that are unstable. The FPY sublibraries are used to determine what
fission products are produced when a fission reaction occurs. This information
is then aggregated and written to an XML file that serves as the depletion chain
file that is used at runtime to construct a burnup matrix.

@@

Some nuclides that have fission reactions in ENDF/B-VII.1, such as $^{243}$Pu,
do not have corresponding FPYs. For these nuclides, a decision must be made on
what set of FPYs to use when fission occurs. This issue was briefly studied by
\citet{fensin2015pne}, who found that the proportion of fission reactions that
occur in these nuclides is usually small enough that the choice of FPYs is not
of great importance. For the depletion chains used in OpenMC, we have adopted
the method used in Serpent, which is as follows:
\begin{enumerate}
  \item If the nuclide missing FPYs is a metastable state, use FPYs from the
  ground state;
  \item Use FPYs from an isotone if available;
  \item If all else fails, use FPYs from $^{235}$U.
\end{enumerate}

@@

Like capture branching ratios, FPYs also depend on the incident neutron energy.
ENDF FPY sublibraries typically include yields tabulated at two or three
energies (thermal, fast, and fusion energies). Complicating matters is the fact
that while these data are specified for a single energy, they actually represent
the observed FPYs for a spectrum of energies (characterized by a single point).
For example, FPYs are often given at \SI{0.0253}{\electronvolt}, but this
actually represents the observed FPYs under a thermal neutron spectrum that
contains a continuum of energies. The ENDF-6 formats manual~\citep{trkov2018bnl}
recommends using linear-linear interpolation between energy points. However, at
the end of a transport solve, we typically have a one-group or multigroup
fission reaction rate---it's not clear how one should use this information to
calculate an effective FPY.

@@

OpenMC's core particle transport solver is written in C++ and is built as a
single executable. Most of the user-facing code, however, is written in Python.
Generally, to run OpenMC, a user writes a Python script that utilizes classes
and functions defined as part of an API to generate XML input files that are
then read by the executable. On top of this, OpenMC is also built as a shared
library with a C++ API that can be called by external programs. OpenMC's Python
API includes bindings to the C++ API; this enables a single run to be broken up
into individual components. For example, a function call from the Python
bindings to the C++ API can initialize the solver (reading inputs, loading cross
sections, etc.) without actually performing a simulation. A separate function
call can run the transport simulation. Further function calls allow a power user
to obtain information about tally results directly in memory from the C++ API.

@@

The abstract \texttt{Integrator} class represents a time integration method.
This class effectively orchestrates the coupled transport--depletion simulation.
Subclasses of \texttt{Integrator} are required to implement a method that takes
an initial vector of nuclide densities, reaction rates returned from an
operator, a time interval, and a power, and returns the resulting nuclide
densities at the end of the step. Thus, \texttt{Integrator} utilizes
\texttt{TransportOperator} in order to obtain reaction rates given $\vect{n}$,
and once the burnup matrix is constructed, it utilizes \texttt{DepSystemSolver}
to deplete with a constant $\vect{A}$ as many times as is necessary. One example
of a subclass of \texttt{Integrator} is \texttt{CECMIntegrator}, which
implements the CE/CM method.

@@

Five coupled transport--depletion simulations were carried out with OpenMC to test
a variety of combinations of depletion chains and methods for handling FPY
energy dependence:
\begin{enumerate}
  \item Full depletion chain, thermal FPYs
  \item Full depletion chain, interpolated FPYs
  \item Reduced depletion chain, thermal FPYs
  \item CASL depletion chain, thermal FPYs
  %\item CASL depletion chain, FPYs based on cutoff method
  \item CASL depletion chain, interpolated FPYs
\end{enumerate}
Because Serpent doesn't have an option for changing the depletion chain or the
method for handling FPY energy dependence, a single coupled transport--depletion
simulation with Serpent was carried out. For all simulations, 20 inactive
batches, 200 active batches, and \num{4.0e6} particles per batch were used,
which resulted in an uncertainty on the effective multiplication factor of about
3--4 pcm.

@@

Four coupled transport--depletion simulations were carried out with OpenMC for
the SFR assembly model:
\begin{enumerate}
  \item Full depletion chain, interpolated FPYs, probability tables on
  \item CASL depletion chain, interpolated FPYs, probability tables on
  \item Full depletion chain, interpolated FPYs, probability tables off
  \item CASL depletion chain, interpolated FPYs, probability tables off
\end{enumerate}
The results for the PWR pincell model indicated that the reduced depletion chain
results in identical behavior to the full depletion chain, so we have not
included a case for the SFR assembly using the reduced depletion chain. Two
coupled transport--depletion simulation with Serpent was carried out, one with
probabilty tables turned on and one with them turned off. For all simulations,
20 inactive batches, 200 active batches, and \num{4.0e6} particles per batch
were used, which resulted in an uncertainty on the effective multiplication
factor of about 3--4 pcm.

@@

A new depletion solver has been implemented in the OpenMC Monte Carlo particle
transport code. The depletion solver is implemented primarily in Python 3 and
interfaces with OpenMC's transport solver through Python bindings to an
underlying C++ API, which enables data transfers to happen in memory rather than
through filesystem I/O. The design of the depletion solver relies on a set of
abstract classes that provide modularity and separation between difference
components, including the time integration methods, the matrix exponential
solver, and the transport operator. In principle, this modularity would allow
the depletion system to be used with another transport code. Rather than relying
on a single time integration method, OpenMC implements several time integration
schemes including the predictor method, standard predictor--corrector methods,
stochastic implicit Euler integrators, and several novel
methods~\citep{josey2017phd, josey2017jcp}. To evaluate the matrix exponential,
OpenMC relies on the incomplete partial fraction form of CRAM by default.

@@

Discuss limitations (not validated against experiment, only as good as data).
Improvements in runtime are needed. Only tested for fission applications.
Further development and testing needed for other application areas, such as
transmutation in fusion reactors.
